A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-10-13 inst/exponential_BT.Rmd
simple cases

>>>>

# Some simple cases

We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
  with the rank of the competitor, the best being rank 1].  Suppose
the focal competitor came first, the likelihood function would be

\begin{equation}
  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
\end{equation}

```{r plot123first}
worker_123first <- function(...){
f <- function(x,r){ x^r/(x + x^2 + x^3) }

x <- seq(from=0.01 , to=1, by=0.01)
plot(c(-3,0), c(-3,0), type='n',
     xlab="log x", ylab="log-likelihood",main="(a)")
points(log(x),log(f(x,1)), type="l", col="black")
points(log(x),log(f(x,2)), type="l", col="red"  )
points(log(x),log(f(x,3)), type="l", col="blue" )

abline(h=0, lty=2)
abline(h = -2)
e <- exp(1)
r2reject <- ((e^2-1)-sqrt(e^4-2*e^2-3))/2
segments(x0=log(r2reject), y0=-2.2,y1=-1.8)

r3reject <- (1+sqrt(1+4*(e^2-1)))/(2*(e^2-1))
segments(x0=log(r3reject), y0=-2.2,y1=-1.8)
legend(x= -3, y= - 0.5, lty=1, col=c("black","red","blue"),
       legend=c("r=1","r=2","r=3"))
}
e <- exp(1)
r2reject <- ((e^2-1)-sqrt(e^4-2*e^2-3))/2
print(r2reject)
r3reject <- (1+sqrt(1+4*(e^2-1)))/(2*(e^2-1))
print(r3reject)

worker_123first()
pdf(file="123first.pdf")
worker_123first()
dev.off()

```

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to different order statistics.
$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
example, if $r=1$ we would have two finishing orders that are
consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
2$.  These are disjoint so the likelihood will be proportional to

$$
\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
$$

After simplification:

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

```{r plot123second}
worker_123second <- function(...){
like <- function(x,r){
  S <- x + x^2 + x^3
  switch(r,
         '1' = x^2/S*x^1/(x^1+x^3) + x^3/S*x^1/(x^1+x^2), # 213+312
         '2' = x^1/S*x^2/(x^2+x^3) + x^3/S*x^2/(x^1+x^2), # 123+321
         '3' = x^1/S*x^3/(x^2+x^3) + x^2/S*x^3/(x^1+x^3)  # 132+231
  )
}

x <- seq(from=0.01, to=1, by=0.01)
plot(c(-3,0),c(-3,0), type='n',
     xlab="log x", ylab="log-likelihood", main="(b)")
points(log(x),log(like(x,1)), type="l", col="black")
points(log(x),log(like(x,2)), type="l", col="red"  )
points(log(x),log(like(x,3)), type="l", col="blue" )

abline(h=0,lty=2)
abline(h=-2)

legend(x = -3, y = -0.5, lty=1, col=c("black","red","blue"),
       legend=c("r=1", "r=2", "r=3"))
}
worker_123second()
pdf(file="123second.pdf")
worker_123second()
dev.off()
```

Now the focal competitor comes last:

```{r plot123third}
worker_123third <- function(...){
like <- function(x,r){
  S <- x + x^2 + x^3
  switch(r,
         '1' = x^3/S*x^2/(x^1+x^2) + x^2/S*x^3/(x^1+x^3), # 213+312
         '2' = x^1/S*x^3/(x^2+x^3) + x^3/S*x^1/(x^1+x^2), # 123+321
         '3' = x^1/S*x^2/(x^2+x^3) + x^2/S*x^1/(x^1+x^3)  # 132+231
  )
}

x <- seq(from=0.01, to=1, by=0.01)
plot(c(-3,0),c(-3,0),type='n',
     xlab="log x", ylab="log-likelihood", main="(c)")
points(log(x),log(like(x,1)), type="l", col="black")
points(log(x),log(like(x,2)), type="l", col="red"  )
points(log(x),log(like(x,3)), type="l", col="blue" )

abline(h=0,lty=2)
abline(h=-2)

legend(x = -3, y = -0.5, lty=1, col=c("black", "red", "blue"),
       legend=c("r=1", "r=2", "r=3"))
}

worker_123third()
pdf(file="123third.pdf")
worker_123third()
dev.off()
```

```{r allthree}
worker_allthree <- function(...){
op <- par(mfrow=c(1,3))
worker_123first()
worker_123second()
worker_123third()
dev.off()
par(op)
}
worker_allthree()
pdf(file="123all.pdf",height=4)
worker_allthree()
dev.off()
```
The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have $1$

### 2025-10-13 inst/hankin-exponential_BT.tex
simple cases

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to $\alpha x^i$ or $\alpha x^j$
coming first where $i,j\neq r$.  For example, if $r=1$ we would have
two finishing orders that are consistent with the observation: $2\succ
1\succ 3$ and $3\succ 1\succ 2$.  These are disjoint so the likelihood
will be proportional to

@@

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have $1

>>>>

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to different order statistics.
$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
example, if $r=1$ we would have two finishing orders that are
consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
2$.  These are disjoint so the likelihood will be proportional to

@@

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x^1}{x^1+x^2+x^3}\left(\frac{x^2}{x^1+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=1\\
    \frac{x^2}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=2\\
    \frac{x^3}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^2}{x^1+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{123all}  % 123all.pdf produced by exponential_BT.Rmd
\caption{Log-likelihood functions \label{123all} for support
  $\in\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}$}
\end{centering}
\end{figure}

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have $1$
